---
title: "Intro to Causal Inference HW1"
author: "Julian McClellan"
date: "February 1, 2018"
output: pdf_document
---

```{r data_read_transform, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(haven) # Read DTA
# dat <- read_dta("K_ELL_T2LEARN_exercise.dta")
og_dat <- read_csv("GSS83_91.csv")

dat <- read_csv("GSS83_91.csv") %>%
  filter(!(tvhours %in% c(-1, 98, 88)),
         !(wrkstat %in% c(0, 9)),
         attend != 9,
         (wrkstat != 0 & wrkstat < 8),
         age >= 25,
         age < 65,
         (tvhours != -1 & tvhours < 98),
         attend < 9,
         educ < 97
         ) %>%
  mutate(wd = ifelse(wrkstat >= 3 & wrkstat <= 8, 1, 0),
         race_d = ifelse(race == 2, 1, 0),
         atddmy = ifelse(attend >= 4 & attend <= 8, 1, 0),
         educ5 = educ, 
         educ5 = case_when(
           (educ5 >= 0 & educ5 <= 11) ~ 1,
           educ5 == 12 ~ 2,
           (educ5 >= 13 & educ5 <= 15) ~ 3,
           educ5 == 16 ~ 4,
           (educ5 >= 17 & educ5 <= 20) ~ 5),
         age10c = floor((age - 5) / 10),
         atddmy = factor(atddmy, levels = c(0, 1), labels = c(
           "Attends church less than once a month",
           "Attends church at least once a month"
         )),
         
         # INteraction
         int = wd * ifelse((age10c >= 3 & age10c <= 4), 1, 0),
         educ5 = factor(educ5, levels = c(2, 1, 3, 4, 5)),
         age10c = factor(age10c, levels = c(2, 3, 4, 5))
         )


# Make variables necessary
```

### 1.
Run the logistic regression to predict the treatment variable ATDDMY including AGE10C, EDUC5, RACE_D, WD, and INT as covariates. Treat AGE10C and EDUC5 as categorical variables by using the first category (25-34) as the baseline category for age, and the second category (high school graduation) as the baseline category for education.

```{r 1}
q_1_logit <- glm(atddmy ~ age10c + educ5 + race_d + wd + int, data = dat, family = binomial)

odds <- exp(coef(q_1_logit))
```

### 2.
Describe the effects of (1) race dummy variable, (2) the effect of employment status for (a) ages 35-54, and (b) other ages on the odds of attending churches at least once a month.

##### (1)
We exponentiatate the coefficients of the logistic regression results. For the race dummy (`RACE_D`) variable, we find that it doubles the odds (`r odds[race_d]`) of attending churches at least once a month.

##### (2)
###### (a)
For the employment status variable (`INT`) we find that it nearly halves the odds of attending church at least once a month (`r odds["int"]`).

###### (b)
For the effect of employment status (`WD`) we find that it increases the odds of attending church at least once a month by ~ 44% (`r odds["wd"]`)


### 3. Using the cross-classification of `LGT_P10` (which is ten times of the logit of the propensity score rounded to an integer value) and `ATTDDMY`, identify (a) the area of common support by the values of the `LGT-P10`, and (b) the number of sample subjects outside the common area of support.

```{r 3}
library(modelr)
logit2prob <- function(x){
  (exp(x) / (1 + exp(x)))
}

dat <- dat %>%
  add_predictions(q_1_logit) %>%
  mutate(prob = logit2prob(pred),
         lgt_p = log(prob / (1 - prob)),
         lgt_p10 = floor(lgt_p * 10 + .5))


dat %>%
  ggplot(aes(x = lgt_p10)) +
  geom_histogram(bins = 10) + 
  facet_wrap(~atddmy) +
  xlab("LGT-P10")


support_table <- dat %>%
  group_by(atddmy) %>%
  summarise(count = n(), min_lgt_p10 = min(lgt_p10), max_lgt_p10 = max(lgt_p10))


library(knitr)
kable(support_table)
```
We see that the common support, using the values of `LGT-P10` includes `r
support_table$count[1] + support_table$count[2]` sample subjects, while the number
of sample subjects outside of the common area of support is 0.

#### 4. 
Define strata as follows using the `LGT_P10`.
(1) [-5, 3], (2) [-2, -1], (3) 0, (4) 1, (5) 2, (6) 3, (7) [4-6], (8) [7, 9], (9) [10, 14]

```{r 4}
dat <- dat %>%
  mutate(strata = case_when(
           (lgt_p10 >= -5 & lgt_p10 <= -3) ~ 1,
           (lgt_p10 >= -2 & lgt_p10 <= -1) ~ 2,
           (lgt_p10 == 0) ~ 3,
           (lgt_p10 == 1) ~ 4,
           lgt_p10 == 2 ~ 5,
           lgt_p10 == 3 ~ 6,
           (lgt_p10 >= 4 & lgt_p10 <= 6) ~ 7,
           (lgt_p10 >= 7 & lgt_p10 <= 9) ~ 8,
           (lgt_p10 >= 10 & lgt_p10 <= 14) ~ 9)) %>%
  na.omit() # Take off values with LGT_P10 outside of given range
```

#### 5.
Calculate the mean of the logit of propensity score (`LOGIT_P`) by strata for each category of the treatment variable.

```{r 5}
q_5_table <- dat %>%
  group_by(strata, atddmy) %>%
  summarise(mean_logit_prop = mean(lgt_p))

strata_groups <- seq(1, 9) %>% 
  map(~ filter(dat, strata == .)) %>%
  map_dbl(~ .$p.value)

strata_groups_2 <- dat %>%
  t.test(lgt_p ~ atddmy, data = .)


kable(q_5_table)
```

#### 6. Test the significance of (1) the difference in the logit between the treatment group and the control group for each stratum, and (2) estimate the average difference in the logit weighted by the proportion of samples in each stratum, and its standard error. Confirm that no significant differences in the logit exist.


##### (1)



#### 7. Test the conditional independence between `ATDDMY` and each of the four variables (`EDUC5`, `RACE_D`, `AGE10C`, and `WD`) controlling for strata.

```{r 7}
library(bnlearn)
library(rlang)

ci_vars <- "educ5" # , "race_d", "age10c", "wd")

ci_results <- ci_vars %>%
  map(~ mantelhaen.test(dat$atddmy, pull(dat, .), dat$strata)) %>%
  map_dbl(.$p.value)
```
