---
title: "Intro to Causal Inference HW1"
author: "Julian McClellan"
date: "February 13, 2018"
output: pdf_document
---

```{r data_read_transform, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(haven)
theme_set(theme_minimal())
dat <- read_dta("WEBSTER.dta")
```

### I.1

#### Do you observe any evidence of non-compliance among first graders?

```{r noncomp}
(dat %>%
  group_by(Z, D) %>%
  summarise(n = n()) %>%
  mutate(assignment_group_prop = n / sum(n))
 )
```

Looking at the above table there does seem to be evidence of non-compliance. Looking at the middle 2 rows, 
we see that about ~8% of those kindergartners *not* assigned to a smaller class size (`Z=0`) actually have a smaller class size in first grade (`D=1`).

Similarly, also around ~8% of those kindergartners who *were* assigned a smaller class size (`Z=1`) actually do not have a smaller class size in first grade (`D=0`).

#### If we assume that the monotonicity condition holds, what proportion of students can be viewed as always takers of small class size in first grade among students initially assigned at random to regular classes?

```{r always_takers}
(dat %>%
   filter(Z == 0) %>% # Not assigned 
   group_by(Z, D) %>%
   summarise(n = n()) %>%
   mutate(assignment_group_prop = n / sum(n))
   )
```
We see there are `r 2867 + 248` students initially assigned to regular classes (`Z=0`). `248` of them, or approximately 8% of these students can be viewed as "always takers" because they are in smaller class sizes in first grade (`D=1`).

#### What proportion of students can be viewed as never takers of small class size in first grade among those initially assigned at random to small classes?

```{r never_takers}
(dat %>%
   filter(Z == 1) %>% # Assigned
   group_by(Z, D) %>%
   summarise(n = n()) %>%
   mutate(assignment_group_prop = n / sum(n))
   )
```

We see there are `r 108 + 1293` students initially assigned to small class sizes (`Z=1`). `108` of them, or approximately 8% of these students can be viewed as "never takers" because they did not stay in smaller class sizes in first grade (`D=0`).

#### How did the average first-grade reading achievement of the never-takers compare with those assigned at random to the regular classes?

```{r avg_rd_lvl}
nt_table_slice <- dat %>%
  group_by(Z, D) %>%
  summarise(n = n(),
            mean_y = mean(Y),
            Z1 = "hi")

reg_slice <- dat %>%
  filter(Z == 0) %>%
  group_by(Z) %>%
  summarise(n = n(), 
            mean_y = mean(Y),
            Z1 = "hi")

(new_tab <- bind_rows(nt_table_slice[3, c(1, 3, 4, 5)], reg_slice) %>%
    mutate(Z1 = ifelse(Z == 1, "Never Takers", "Regular Classes")) %>%
    rename(`Never takers or Regular Class Assignment` = Z1) %>%
    dplyr::select(-Z)
  ) 
```

### I.2

#### (1) The exogeneity of Z

```{r exo_endo_test}
library(AER)
# Will be used later
filtered_dat <- dat %>%
                # We keep only those who complied with the random assignment
                # (Z = 0 & D = 0) | (Z = 1 & D = 1)
                filter(
                  (Z == 1 & D == 1) | (Z == 0 & D == 0)
                )
iv2 <-  ivreg(Y ~ D
                | Z, data = dat)

(summary(iv2, vcov = sandwich, diagnostics = T))
```
The exogeneity of `Z` seems plausible since we fail to reject (at 95%
confidence) with the Wu-Hausman test.
 
 
#### (2) Exclusion Assumption

#### (3) Monotonicity assumption

#### (4) Correlated effects


### I.3

```{r}
lmi3 <- lm(Y ~ D, data=dat)
```

#### What is the average mean difference in first-grade reading achievement between those attending small classes (`D=1`) and those attending regular classes (`D=0`) in first grade? What is the effect size?

```{r}
(summary(lmi3))
```
Looking at the above summary table for our regression of `Y` on `D` the average mean difference appears to be `r coef(summary(lmi3))[2]`. However, this the p-value is only `r coef(summary(lmi3))[2,4]`.

```{r itt_effect}
alpha_1 <- lm(D ~ Z, data=dat)
veta_1 <- lm(Y ~ Z, data=dat)
```

### I.4


#### Regression Models

$$Y = \beta_0 + \beta_1D + \epsilon$$

However, `D` is correlated with $\epsilon$, so $E[\hat{\beta_1}]\neq \beta_1$. (I.e. $\hat{\beta_1}$ is a biased estimator of $\beta_1$)

Thus we utilize 2SLS. Stage 1:

$$D = \gamma_0 + \gamma_1Z + u$$
$$\hat{D} = \hat{\gamma_0} + \hat{\gamma_1}Z + u$$

Because we assume that `Z` is independent of $\epsilon$, thus $\hat{D}$ is independent of $epsilon$ as well. Now we move on to Stage 2:

$$D = \beta_0 + \beta_1\hat{D} + \epsilon$$

Now $\hat{\beta_1}$ is an unbiased estimate of $\beta_1$


#### According to your analytic results, what is the average effect of class size reduction on the reading achievement of students who complied with the random assignment for two consecutive years? Is the effect significant.

```{r 2sls}
iv2_2 <-  ivreg(Y ~ D
                | Z, data = dat)
(summary(iv2_2))
```
The estimated coefficient for `D` is `r coef(iv2)[2]`, and we see that it is significant at the 5% level.


### I.5

#### Could you give a reason why the result from I.4 might be different from the PF estimate obtained ealier when you simply regress `Y` on `D`? Confirm also that the ratio of the estimate of $\gamma_1$ to the estimate of $\alpha_1$ in I.3 is the same (except for a very minor difference) as the 2SLS estimate.
